---
title: "ggplot Demo"
author: "KJF"
date: "March 27, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Script to demo ggplot use of factors 
Load packages and demo Sunapee data
``` {r message=FALSE}
library(tidyverse)
library(lubridate)

raw <- read_csv("LMP_L1daily_temp_merge.csv")

names(raw)
```

Whoa, those column names are out of control!   
We'll have to massage the data into a more user-friendly format: 
```{r}
startDate = ymd('2005-01-01')

manualTemp <- raw %>% 
  filter(date >= startDate) %>% # Retain dates on or after start date
  select(-datasource) %>% # Drop datasource column
  rename_at(vars(starts_with("Temp_")), funs(sub("Temp_", "", .))) %>% # 3 steps 
  rename_at(vars(ends_with("_degC")), funs(sub("m_degC","",.))) %>% # to streamline
  rename_at(vars(contains("p")), funs(sub("p", ".",.))) %>% # column names
  gather(Depth, Temp, "0":"33") %>% # Reshape wide to long
  mutate(Depth = as.double(Depth), Temp = round(as.double(Temp), 2)) %>% # Parse Depth and Temp
  na.omit() # Remove rows with NA values
  
head(manualTemp)
```
Much better!  
Note that while we specificed Depth and temp as being double, loc is currently parsed as a character.

If you make a plot of temperature over time with depths as different colors, the depths default to a color gradient, since their structure is "double":
```{r}
ggplot(manualTemp, aes(x = date, y = Temp, col = Depth)) + 
  geom_point() +
  scale_y_continuous("Water Temperature (C)", limits= c(-5,30)) +
  scale_x_date("Year", date_breaks= "1 year", date_labels= "%Y")
```

If, instead, you want discrete colors, you would set Depth as a factor. You can do this in the ggplot command directly.
```{r}
ggplot(manualTemp, aes(x = date, y = Temp, col = factor(Depth))) + 
  geom_point() +
  scale_y_continuous("Water Temperature (C)", limits= c(-5,30)) +
  scale_x_date("Year", date_breaks= "1 year", date_labels= "%Y")
```

Note that for these plots, you're lumping all the locations together. Maybe you just want to look at 210 and the buoy at 1, 10, and 20 meters.
```{r}
manTemp <- manualTemp %>%
  filter((loc == '210' | loc=='buoy') & (Depth %in% c(1, 10, 20)))
```

Now we'll plot with each location as a shape and each depth as a color:
```{r}
ggplot(manTemp, aes(x = date, y = Temp, pch = loc, col = Depth)) + 
  geom_point() +
  scale_y_continuous("Water Temperature (C)", limits= c(-5,30)) +
  scale_x_date("Year", date_breaks= "1 year", date_labels= "%Y")
```  
  
Here, it's disingenuous to have Depth as a color gradient, because we're only looking at 3 discrete depths. Thus we need to have it as a factor!  
We could add factor() to the aes in the plotting command, as we did above, or recode Depth as a factor in the tibble itself: 
```{r}
manTemp <- manTemp %>%
  mutate(Depth = factor(Depth))

head(manTemp)

ggplot(manTemp, aes(x = date, y = Temp, pch = loc, col = Depth)) + 
  geom_point() +
  scale_y_continuous("Water Temperature (C)", limits= c(-5, 30)) +
  scale_x_date("Year", date_breaks= "1 year", date_labels= "%Y")
```

Notice how instead of using subset commands to select each location or depth, you add them in to the aes command, and they're automatically added to the legend. Whoo!